Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-06-21 docs/paper.Rmd
write theoretical background (unified growth models)

Finally, I discuss more recent refinements that enrich the basic model by incorporating credit constraints \citep{doepke2004accounting}, intra-household bargaining \citep{doepke2019bargaining}, and multi-dimensional child quality \citep{hoddinott2013economic, kalemli2002does}, with a special emphasis on health and nutrition. My focus is on how the Q–Q framework has been applied to understand fertility and child investment patterns, especially in developing country contexts where resource constraints and health outcomes are paramount.

@@

\subsubsection{Becker’s Static Model}
Becker’s early work introduced an economic approach to fertility, treating children analogous to “durable goods” that yield utility to parents but come at a cost \citep{doepke2015gary}. In Becker’s 1960 model, a household derives satisfaction from the number of children ($n$) and from the quality of each child ($q$), alongside conventional consumption of other goods ($y$). A simple representation is a utility function:

@@

with $U$ increasing in each argument up to some satiation point. Here quality $q$ can be thought of as the expenditure or investment per child (e.g. education spending, health care, nutrition), assumed for now to be the same for each child. Parents face a budget constraint that links quantity and quality: raising more children dilutes the resources available per child. A prototypical budget constraint (in static form) can be written as:

@@

>>>>

Finally, I discuss more recent refinements that enrich the basic model by incorporating credit constraints \citep{doepke2004accounting}, intra-household bargaining \citep{doepke2019bargaining}, and multi-dimensional child quality \citep{hoddinott2013economic, kalemli2002does}, with a special emphasis on health and nutrition. The literature review focuses on how the Q–Q framework has been applied to understand fertility and child investment patterns, especially in developing country contexts where resource constraints and health outcomes are paramount.

@@

\subsubsection{Becker’s Static Model}

Becker’s early work introduced an economic model of fertility and treated children as durable goods that provide utility to parents but impose costs \citep{doepke2015gary}. In Becker’s 1960 model, a household derives satisfaction from the number of children ($n$) and from the quality of each child ($q$), alongside conventional consumption of other goods ($y$). A simple representation is a utility function:

@@

with $U$ increasing in each argument up to some satiation point. Here, quality $q$ can be thought of as the expenditure or investment per child (e.g. education spending, health care, nutrition), assumed for now to be the same for each child. Parents face a budget constraint that links quantity and quality: raising more children dilutes the resources available per child. A prototypical budget constraint (in static form) can be written as:

@@

\subsubsection{Unified Growth Models}

The unified growth theory, developed in the late 1990s and 2000s (notably by Oded Galor and co-authors), seeks to explain in one framework the entire sweep of economic development – from Malthusian stagnation, through the demographic transition, to modern growth. A central puzzle it addresses is why fertility rates, which were historically high and invariant to income in the Malthusian era, began to decline sharply in tandem with industrialization and rising incomes, eventually stabilizing at much lower levels in developed economies. The Q–Q trade-off provides a key part of the answer in these models. \citet{galor2000population} and \citet{galor2002natural} explicitly incorporate parental choices about the quantity and quality of children and show how changes in the economic environment alter those choices and trigger demographic transitions.

In \citet{galor2000population}’s model, for instance, technological progress gradually increases the return to human capital, especially in skilled occupations. In the early stages, when production relies on basic tools and techniques, unskilled labor holds more value. Under these conditions, parents have little reason to invest in formal schooling. Children are expected to contribute economically through agricultural work, domestic tasks, or low-skill jobs in workshops and factories. Fertility remains high because children impose a low financial burden and generate immediate returns. As technology becomes more advanced—such as during the Industrial Revolution—the earnings gap between skilled and unskilled labor widens. Education begins to offer significant advantages in the labor market. In response, parents adjust by having fewer children and placing greater emphasis on each child’s development, including school attendance and better health care.

Evidently, industrialization raises the economic value of skilled labor, which alters household incentives. As returns to education increase, parents begin to favor investments in child quality over child quantity. This shift results in declining fertility because families choose to have fewer children and allocate more resources to each. The feedback effect is significant: higher educational investment raises productivity in the next generation, which in turn accelerates technological advancement and further increases the returns to human capital. Over time, the economy moves from a state of high fertility and low growth to one characterized by low fertility and sustained growth.

Furthermore, \citet{galor2002natural} introduced an evolutionary refinement to the unified growth framework by accounting for heterogeneity in parental preferences. During the Malthusian period, some families placed greater emphasis on child quality, such as education, while others prioritized quantity. In a stagnant economy with limited returns to education, high-fertility lineages maintained a numerical advantage and suppressed average human capital. As technological progress increased the returns to education, families that valued quality gained an economic edge. Their children acquired more human capital and achieved higher income and survival rates. These advantages allowed such families to grow in relative size. Over time, this process resembled a form of evolutionary selection, gradually favoring quality-oriented parental types and shifting the population toward greater emphasis on child human capital. These dynamics strengthened the shift from high-fertility, low-education regimes to low-fertility, high-investment family structures.
In formal overlapping-generations models, \citet{galor2002natural} demonstrate that this evolutionary adaptation accelerates the demographic transition. Their framework accounts for the rapid and widespread drop in fertility once it takes effect. Higher returns to human capital push parents to favor quality, while preferences for quality begin to dominate within the population. These forces support the emergence of a low-fertility, high-investment equilibrium and establish a unified explanation for both economic development and demographic change.

Importantly, unified growth models identify several complementary mechanisms that reinforce the basic Q–Q trade-off during development. One is the decline of child labor. As the economy modernizes, the value of child labor falls, both because legal reforms often restrict child labor and because parents realize the earnings their children could make as unskilled laborers are paltry compared to the potential returns if those children instead spend time in school. \citet{hazan2002child} formally show that when child labor becomes less profitable relative to adult (skilled) labor, parents further reduce fertility and invest more in each child’s education. Historical evidence from England, for instance, indicates that during industrialization the wages of children (relative to adults) dropped significantly, especially in skilled families, and this was accompanied by parents pulling children out of work to send them to school. \citet{galor2006human} even argue that capitalist industrialists supported public education laws and child labor bans as a way to increase the human capital of the workforce, inadvertently hastening the fertility transition.

Another mechanism is the rise in life expectancy and child survival. Improvements in sanitation, nutrition, and medical knowledge in developing societies led to more children surviving to adulthood. While the earliest unified growth models treated mortality as exogenous or ignored it, later research demonstrated that declining child mortality can trigger lower fertility as well– parents no longer need “extra” births for insurance once they are confident their existing children will survive. In other words, increased child survival and the quality–quantity trade-off are complementary explanations for fertility decline that operate in tandem. When fewer births are lost to disease, parents can achieve a desired number of surviving offspring with fewer total births, and they tend to reallocate effort into each child’s health and education.

The overall effect is a reinforcing cycle: better health raises the returns to schooling (healthier children can learn more effectively and have longer working lives), which further encourages educational investments and reduces fertility. Indeed, Galor notes that human capital should be interpreted broadly to include health as well as schooling; in unified growth theory, improvements in nutrition and physical well-being were crucial to making labor more productive and thus were part and parcel of the rise in demand for human capital.

The unified growth literature places the Q–Q model within a more general account of economic and demographic change. In this framework, higher income or stronger returns to child quality reduce fertility and help shift economies from stagnation toward sustained growth. Several mechanisms support this transition, such as a fall in child labor, a drop in child mortality, and a shift in parental priorities. These models explain not only the presence of a quantity–quality trade-off but also its rising influence at a specific point in history. The evidence supports these claims: countries that saw earlier increases in returns to education experienced earlier fertility decline, while delays in reforms, such as public education or health access, corresponded to prolonged high fertility. As a result, the Q–Q mechanism forms a key component of unified growth theory.

### 2025-06-21 docs/paper.tex
write theoretical background (unified growth models)

Finally, I discuss more recent refinements that enrich the basic model
by incorporating credit constraints \citep{doepke2004accounting},
intra-household bargaining \citep{doepke2019bargaining}, and
multi-dimensional child quality
\citep{hoddinott2013economic, kalemli2002does}, with a special emphasis
on health and nutrition. My focus is on how the Q--Q framework has been
applied to understand fertility and child investment patterns,
especially in developing country contexts where resource constraints and
health outcomes are paramount.

@@

Becker's early work introduced an economic approach to fertility,
treating children analogous to ``durable goods'' that yield utility to
parents but come at a cost \citep{doepke2015gary}. In Becker's 1960
model, a household derives satisfaction from the number of children
(\(n\)) and from the quality of each child (\(q\)), alongside
conventional consumption of other goods (\(y\)). A simple representation
is a utility function:

@@

with \(U\) increasing in each argument up to some satiation point. Here
quality \(q\) can be thought of as the expenditure or investment per
child (e.g.~education spending, health care, nutrition), assumed for now
to be the same for each child. Parents face a budget constraint that
links quantity and quality: raising more children dilutes the resources
available per child. A prototypical budget constraint (in static form)
can be written as:

@@

@@

>>>>

Finally, I discuss more recent refinements that enrich the basic model
by incorporating credit constraints \citep{doepke2004accounting},
intra-household bargaining \citep{doepke2019bargaining}, and
multi-dimensional child quality
\citep{hoddinott2013economic, kalemli2002does}, with a special emphasis
on health and nutrition. The literature review focuses on how the Q--Q
framework has been applied to understand fertility and child investment
patterns, especially in developing country contexts where resource
constraints and health outcomes are paramount.

@@

\subsubsection{Becker’s Static Model}

Becker's early work introduced an economic model of fertility and
treated children as durable goods that provide utility to parents but
impose costs \citep{doepke2015gary}. In Becker's 1960 model, a household
derives satisfaction from the number of children (\(n\)) and from the
quality of each child (\(q\)), alongside conventional consumption of
other goods (\(y\)). A simple representation is a utility function:

@@

with \(U\) increasing in each argument up to some satiation point. Here,
quality \(q\) can be thought of as the expenditure or investment per
child (e.g.~education spending, health care, nutrition), assumed for now
to be the same for each child. Parents face a budget constraint that
links quantity and quality: raising more children dilutes the resources
available per child. A prototypical budget constraint (in static form)
can be written as:

@@

\subsubsection{Intergenerational Models: Altruism and Child Endowments}

@@

\subsubsection{Unified Growth Models}

The unified growth theory, developed in the late 1990s and 2000s
(notably by Oded Galor and co-authors), seeks to explain in one
framework the entire sweep of economic development -- from Malthusian
stagnation, through the demographic transition, to modern growth. A
central puzzle it addresses is why fertility rates, which were
historically high and invariant to income in the Malthusian era, began
to decline sharply in tandem with industrialization and rising incomes,
eventually stabilizing at much lower levels in developed economies. The
Q--Q trade-off provides a key part of the answer in these models.
\citet{galor2000population} and \citet{galor2002natural} explicitly
incorporate parental choices about the quantity and quality of children
and show how changes in the economic environment alter those choices and
trigger demographic transitions.

In \citet{galor2000population}'s model, for instance, technological
progress gradually increases the return to human capital, especially in
skilled occupations. In the early stages, when production relies on
basic tools and techniques, unskilled labor holds more value. Under
these conditions, parents have little reason to invest in formal
schooling. Children are expected to contribute economically through
agricultural work, domestic tasks, or low-skill jobs in workshops and
factories. Fertility remains high because children impose a low
financial burden and generate immediate returns. As technology becomes
more advanced---such as during the Industrial Revolution---the earnings
gap between skilled and unskilled labor widens. Education begins to
offer significant advantages in the labor market. In response, parents
adjust by having fewer children and placing greater emphasis on each
child's development, including school attendance and better health care.

Evidently, industrialization raises the economic value of skilled labor,
which alters household incentives. As returns to education increase,
parents begin to favor investments in child quality over child quantity.
This shift results in declining fertility because families choose to
have fewer children and allocate more resources to each. The feedback
effect is significant: higher educational investment raises productivity
in the next generation, which in turn accelerates technological
advancement and further increases the returns to human capital. Over
time, the economy moves from a state of high fertility and low growth to
one characterized by low fertility and sustained growth.

Furthermore, \citet{galor2002natural} introduced an evolutionary
refinement to the unified growth framework by accounting for
heterogeneity in parental preferences. During the Malthusian period,
some families placed greater emphasis on child quality, such as
education, while others prioritized quantity. In a stagnant economy with
limited returns to education, high-fertility lineages maintained a
numerical advantage and suppressed average human capital. As
technological progress increased the returns to education, families that
valued quality gained an economic edge. Their children acquired more
human capital and achieved higher income and survival rates. These
advantages allowed such families to grow in relative size. Over time,
this process resembled a form of evolutionary selection, gradually
favoring quality-oriented parental types and shifting the population
toward greater emphasis on child human capital. These dynamics
strengthened the shift from high-fertility, low-education regimes to
low-fertility, high-investment family structures. In formal
overlapping-generations models, \citet{galor2002natural} demonstrate
that this evolutionary adaptation accelerates the demographic
transition. Their framework accounts for the rapid and widespread drop
in fertility once it takes effect. Higher returns to human capital push
parents to favor quality, while preferences for quality begin to
dominate within the population. These forces support the emergence of a
low-fertility, high-investment equilibrium and establish a unified
explanation for both economic development and demographic change.

Importantly, unified growth models identify several complementary
mechanisms that reinforce the basic Q--Q trade-off during development.
One is the decline of child labor. As the economy modernizes, the value
of child labor falls, both because legal reforms often restrict child
labor and because parents realize the earnings their children could make
as unskilled laborers are paltry compared to the potential returns if
those children instead spend time in school. \citet{hazan2002child}
formally show that when child labor becomes less profitable relative to
adult (skilled) labor, parents further reduce fertility and invest more
in each child's education. Historical evidence from England, for
instance, indicates that during industrialization the wages of children
(relative to adults) dropped significantly, especially in skilled
families, and this was accompanied by parents pulling children out of
work to send them to school. \citet{galor2006human} even argue that
capitalist industrialists supported public education laws and child
labor bans as a way to increase the human capital of the workforce,
inadvertently hastening the fertility transition.

Another mechanism is the rise in life expectancy and child survival.
Improvements in sanitation, nutrition, and medical knowledge in
developing societies led to more children surviving to adulthood. While
the earliest unified growth models treated mortality as exogenous or
ignored it, later research demonstrated that declining child mortality
can trigger lower fertility as well-- parents no longer need ``extra''
births for insurance once they are confident their existing children
will survive. In other words, increased child survival and the
quality--quantity trade-off are complementary explanations for fertility
decline that operate in tandem. When fewer births are lost to disease,
parents can achieve a desired number of surviving offspring with fewer
total births, and they tend to reallocate effort into each child's
health and education.

The overall effect is a reinforcing cycle: better health raises the
returns to schooling (healthier children can learn more effectively and
have longer working lives), which further encourages educational
investments and reduces fertility. Indeed, Galor notes that human
capital should be interpreted broadly to include health as well as
schooling; in unified growth theory, improvements in nutrition and
physical well-being were crucial to making labor more productive and
thus were part and parcel of the rise in demand for human capital.

The unified growth literature places the Q--Q model within a more
general account of economic and demographic change. In this framework,
higher income or stronger returns to child quality reduce fertility and
help shift economies from stagnation toward sustained growth. Several
mechanisms support this transition, such as a fall in child labor, a
drop in child mortality, and a shift in parental priorities. These
models explain not only the presence of a quantity--quality trade-off
but also its rising influence at a specific point in history. The
evidence supports these claims: countries that saw earlier increases in
returns to education experienced earlier fertility decline, while delays
in reforms, such as public education or health access, corresponded to
prolonged high fertility. As a result, the Q--Q mechanism forms a key
component of unified growth theory.
